---
editor_options: 
  chunk_output_type: console
---

```{r setup05, include=FALSE}
options(digits = 3)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
theme_set(theme_bw()+theme(text = element_text(size=18)))
```

# Vowel formants' normalization

This section is based on [@adank03].

There are three possible sources of variation in vowel formants measurements [@ladefoged57; @pols73: 1095; @adank03]:

* acoustic variation;
* speaker variation;
   * sociolinguistic;
   * anatomical/physiological variation;
* and measurement error ("residual variance" in [@pols73]).

There are a lot of researchers aimed to reduce speaker-related variation using **acoustic vowel normalization** (e. g. [@gerstman68; @lobanov71; @syrdal86]). However there are some researches that afraid that normalization procedures can reduce interesting for the linguistics information like sociolinguistic/dialectal signal in data [@hindle78; @disner80; @thomas02 174--175].

Human listeners deal seemingly effortlessly with all three possible sources of variation, but the dataset from [@peterson52] shows extrordinary variation:

```{r, fig.width=12}
library(phonTools)
data(pb52)
pb52 %>% 
  mutate(vowel = ipa::convert_phonetics(vowel, from = "xsampa", to = "ipa")) %>%
  ggplot(aes(f2, f1, label = vowel, color = vowel))+
  stat_ellipse(show.legend = FALSE)+
  geom_text(show.legend = FALSE)+
  scale_x_reverse()+
  scale_y_reverse()+
  coord_fixed()+
  labs(caption = "data from [Peterson, Barney 1952]")
```

## Acoustic vowel normalization procedures

There are several classes of vowel normalization procedures:

* formant-based procedures [@gerstman68; @lobanov71; @fant75; @syrdal86; @miller89];
* whole-spectrum procedures [@klein70; @pols73; @bladon81; @bladon82; @klatt82];
* Neural networks [@weenink93; @weenink97].

Formant-based procedures are the most compact (just 2- or 3-dimensional represetations) and comparable crosslinguisticaly.

In [@adank03] author compared 11 methods of vowel normalization:

|    | abb                  | method                                            |
|----|----------------------|---------------------------------------------------|
| 1  | HZ                   | the baseline condition, formants in Hz            |
| 2  | LOG                  | a log-transformation of the frequency scale       |
| 3  | BARK                 | a bark-transformation of the frequency scale      |
| 4  | MEL                  | a mel-transformation of the frequency scale       |
| 5  | ERB                  | an ERB-transformation of the frequency scale      |
| 6  | GERSTMAN             | Gerstman’s (1968) range normalization             |
| 7  | LOBANOV              | Lobanov’s (1971) z-score transformation           |
| 8  | NORDSTRÖM & LINDBLOM | Nordström & Lindblom’s (1975) vocal-tract scaling |
| 9  | CLIH i4              | Nearey’s (1978) individual log-mean procedure     |
| 10 | CLIH s4              | Nearey’s (1978) shared log-mean procedure         |
| 11 | SYRDAL & GOPAL       | Syrdal & Gopal’s (1986) bark-distance model       |
| 12 | MILLER               | Miller’s (1989) formant-ratio model               |

## [@lobanov71] z-score transformation

The idea behind the Lobanov's method is simple z-normalization. Imagine some random distribution:

```{r, echo=FALSE}
set.seed(42)
fake_data <- tibble(x = rnorm(151, mean = 57, sd = 2))
fake_data %>% 
  ggplot(aes(x))+
  geom_histogram(bins = 15)
```

If we apply the folowing normalization, the distribution form will be the same, however the scale will be unified with mean = 0 and standard deviation = 1:

$$x_{normalized} = \frac{x-\mu}{\sigma}$$


```{r, echo = FALSE}
fake_data %>% 
  mutate(x = scale(x)) %>% 
  ggplot(aes(x))+
  geom_histogram(bins = 15)
```

```{r}
library(phonTools)
data(pb52)
pb52 %>% 
  group_by(speaker) %>% 
  mutate(vowel = ipa::convert_phonetics(vowel, from = "xsampa", to = "ipa"),
         f1 = scale(f1),
         f2 = scale(f2)) %>%
  ggplot(aes(f2, f1, label = vowel, color = vowel))+
  stat_ellipse()+
  geom_text()+
  scale_x_reverse()+
  scale_y_reverse()
```

You can find implementation of other methods in R package `vowels`.

Try to normalize and visualize data from the dataset Hillenbrand et al. (1995), stored in `h95` variable in the package `phonTools`.



